---
title: "Day 6 Comp"
author: "Peter Menzies"
date: "8/16/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(palmerpenguins)
```

## Warm-up:

```{r}
View(us_rent_income)
```

### Tidy it up

```{r}
tidy_income <- us_rent_income %>% 
  pivot_wider(names_from = variable, values_from = estimate) %>% 
  select(-moe)

untidy <- tidy_income %>% 
  pivot_longer(cols = "income", "rent", names_to = "variable", values_to = "estimate")
```

### filter to keep or exclude rows

1. exact macth (single condition)

Only keep observations where the island is Biscoe
```{r}
ex_1 <- filter(.data = penguins, island == "Biscoe")
unique(ex_1$island)
```
2. "and statement within filter

only keep observations if the penguin species is chinstrap and the year is 2008
```{r}
ex_2 <- penguins %>%
  filter(species == "Chinstrap", year == 2008)
```

3. Exclude Gentoos

Keep observations if the penguin species is NOT "Gentoo"
```{r}
ex_3 <- penguins %>% 
  filter(species != "Gentoo")
```

Practice

```{r}
adelie_flipper <- penguins %>% 
  filter(species == "Adelie", flipper_length_mm >= 185)

# Check our ouput:
unique(adelie_flipper$species)
min(adelie_flipper$flipper_length_mm)
```

4. Use the %in% operator

```{r}
practice_2 <- penguins %>% 
  filter(island %in% c("Dream", "Torgersen"))

# == in the above statement would look for "Dream" and "Torgersen" alternatingly in the column

# add an `!` in front of the VARIABLE name to exlude the values in the vector
```

### Select() to select or exlude COLUMNS

5.Keep a single column
```{r}
ex_5 <- penguins %>% select(species) %>% 
  pull(species)

ex_5 <- select(.data = penguins, species)
```

6. Keep multiple columns by name

```{r}
ex_6 <- select(.data = penguins, species, bill_length_mm:flipper_length_mm, year)
```

7. Choose or exclude columns based on postiion
```{r}
ex_7 <- penguins %>% 
  select(-4)
```

8. A Piped sequence where you first: only keep columns for species, body mass, and island, and second: only include body mass is greater than 3600
```{r}
ex_8 <- penguins %>% 
  select(species, body_mass_g, island) %>% 
  filter(body_mass_g > 3600)
```

### Mutate() to add or update existing columns, while keeping the onees already there

9. Add a column that has body mass converted to kg

```{r}
ex_9 <- penguins %>% 
  mutate(body_mass_kg = body_mass_g / 1000) %>% 
  mutate(flipper_length_m = flipper_length_mm / 1000)

```

10. Mutate to update an existing column
```{r}
ex_10 <- penguins %>% 
  mutate(island = str_replace(island, pattern = "sen", replacement = "ooski"))
```

11. Update the class of a column using mutate()
```{r}
# Tried using assignment operator and didn't work
ex_11 <- penguins %>% 
  mutate(bill_depth_mm = as.character(bill_depth_mm))
```

### case-when(): your fav if-else

12. Create a new column that contains "small" if flipper length is less than 185 mm, "medium" otherwise

```{r}
ex_12 <- penguins %>% 
  mutate(binned_size = case_when(
    flipper_length_mm < 185 ~ "small",
    flipper_length_mm >= 185 ~ "medium"
  ))
```

Task: Starting from penguins, create a new col "bird_bill_bin" that contains:

- If bill length < than 38mm: "short bill"
- If bill length >= 38mm & < 43mm: "medium bill"
- If bill length >= 43: "long bill"

```{r}
bill_task <- penguins %>% 
  mutate(bird_bill_bin = case_when(
    bill_length_mm < 38 ~ "short bil",
    between(bill_length_mm, 38, 42) ~ "medium bill",
    bill_length_mm >= 43 ~ "long bill"
  ))
```

13. group_by() to say there are groups within the species column

```{r}
#group_by just lets R know to lump similar values within a variable into groups for some further action to come

ex_13 <- penguins %>% 
  group_by(species, island, year) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE))

```

## Practice 

In a single piped sequence

- Only keep penguins that have a body mass > 3500 g
- Only keep columns for species, flipper_length_mm, and sex
- create a summary table that contains the mean, sd, max, and min val of flipper length by species and sex

```{r}
practice_3 <- penguins %>% 
  filter(body_mass_g > 3500) %>% 
  select(species, flipper_length_mm, sex) %>% 
  group_by(species, sex) %>% 
  summarize(mean_flipper_length = mean(flipper_length_mm, na.rm = TRUE),
            sd_flipper_length = sd(flipper_length_mm, na.rm = TRUE),
            max_flipper_length = max(flipper_length_mm, na.rm = TRUE),
            min_flipper_length = min(flipper_length_mm, na.rm = TRUE))

```

